Panthera: Caching in Distributed Computing Systems

### 2014-05-02 paper/paper.tex

%
% Set the title, author, and date.
%
\title{Panthera: A Study of Caching in Distributed Computing}
\author{Dhaivat Pandya \\ Appleton North High School}
\date{}
%
% The document proper.
%
\begin{document}

@@

% Add an abstract.
\abstract{
The adoption of distributed computing systems has grown massively in the past few years. In particular, Apache Hadoop, which allows developers to create applications that run on a cluster of computers, is currently used throughout academia and industry. However, the Hadoop File System fails to effectively utilize memory and local storage in order to reduce waiting time (i.e. latency). In this project, a system, named Panthera, was developed, which creates a caching system for Hadoop metadata and data. Panthera caches both information about files and the file contents, thereby reducing waiting time in downloading and accessing the files and related information. It runs with an unmodified version of Hadoop, meaning that it can integrate easily into existing architecture. Panthera was tested using both single node and multi-
node setups. The results showed an 90.8\% decrease in waiting time with the Panthera metadata cache, and a 95.7\% decrease with the Panthera data cache. Such drastic decreases in latency can greatly increase the efficiency of existing Hadoop applications and also allow the creation of algorithms that were previously not feasible. Panthera has numerous applications in fields ranging from market research to bioinformatics. All code associated with the project will be open sourced to further development in the area.
}

\section{Introduction}
Distributed computing systems are currently employed across academia and industry, with applications varying from market research for department stores to bioinformatics. In particular, the Hadoop distributed system \cite{hadoop} has seen tremendous growth. Hadoop is an open source version of the revolutionary MapReduce system developed at Google. With it, developers can easily take advantage of large, multi-node clusters to solve computational problems.

Current Hadoop deployments exist in the thousands of nodes at companies such as Cloudera, Facebook, Yahoo, etc. However, especially in larger networks, file access latency can be a significant issue. The Hadoop File System currently does not use local (i.e. on the client) random access memory (RAM) to improve latency.

Latency in distributed systems can have significant effects, and a reduction of the same comes with tremendous benefits. As the RAMCloud project has outlined \cite{ramcloud}, low latency can greatly extend the applications of distributed computing systems. For example, tree traversal algorithms are largely impractical in Hadoop, however, with lower latency, such algorithms may become useful and applicable.

In this paper, we discuss \textit{Panthera}, a cache layer for Hadoop. In Section 2, we discuss the specific challenges associated with such a project. Section 3 details the architecture differences between a standard Hadoop installation and one with \textit{Panthera}.  Sections 4 and 5 detail the differences between metadata and data caching within \textit{Panthera}. Results and conclusions are covered in Sections 6, 7, 8, 9, concluding with a roadmap for future development in Section 10.

\section{Constraints}

There are several systems built on Hadoop that are in widespread use \cite{hbase, cloudbatch, pig}. The Hadoop project is also supported by large corporations with a 
consistent release cycle. Thus, for \textit{Panthera} to be practical, it must operate independently of the existing Hadoop codebase, i.e. as a layer, rather than a modification. This implies that \textit{Panthera} must implemented without a single change in the existing Hadoop codebase, i.e. as a "drop-in" solution.

@@

@@

Figure 1 outlines the architecture differences between a standard Hadoop File System architecture and an installation with Panthera.

@@

Currently, a standard LRU cache is used for each method. Testing of different caching methods and predictive prefetching algorithms is under way.

@@

Tests for the metadata cache were conducted on both single node and multinode setups, whereas the datanode cache testing was run on multinode setups.

@@

The Panthera data cache showed a 95.7\% decrease in the latency in comparison to a standard deviation non-datacached installation. In addition, the standard of deviation was lower by a factor of 27. For any deployed project, maximum levels of latency must be considered when designing a system. Thus, the \textit{status quo} of the non-cached version severely limits possibilities within Hadoop.

@@

Metadata caching has similarly incredible results. \textit{Panthera} showed a 9 fold or 90.8\% improvement in the total latency as measured on the client side in a multinode setup. Additionally, the standard deviation is 24 fold lower in \textit{Panthera}.

@@

\textit{Panthera} fulfils the constraints mentioned at the outset. It is a layer on Hadoop rather than a main repository code modification. Considering the latency measurements, it is clear that it has also met the requirements in terms of benefit. \textit{Panthera} was able to reduce data access latency by 95.7\% and reduce basic metadata access latency by 90.8\%. Additionally, \textit{Panthera} greatly reduced the variation in the access latency, greatly aiding in the pre-development plans and extending the capabilities of distributed computing systems such as Hadoop.

@@

I am also working on a cache-based scheduler for Hadoop. Essentially, it will use information about what files are available in various clients and arrange Hadoop tasks such that the use of the cache is maximized.

>>>>

%
% Set the title, author, and date.
%
\title{Panthera: Caching in Distributed Computing Systems}
\author{Dhaivat Pandya \\ Appleton North High School}
\date{}
%
% The document proper.
%
\begin{document}

@@

% Add an abstract.
\abstract{
	The adoption of distributed computing systems has grown massively in the past few years. In particular, Apache Hadoop, which allows developers to create applications that run on a cluster of computers, is currently used throughout academia and industry. However, the Hadoop File System fails to effectively utilize memory and local storage in order to reduce waiting time (i.e. latency). Thus, there is a very significant wastage of available computational resources. In this project, a system, named Panthera, was developed which reduces waiting time and conserves computational resources by using \textit{caching}, i.e. retaining downloaded files in local memory rather than discarding them, thereby reducing access time. Panthera caches both information about files (i.e. metadata) and the file contents (i.e. data). It runs with an unmodified version of Hadoop, meaning that it can integrate easily into existing architecture. Panthera was tested using both single node and multi-node setups. The results showed a 90.8\% decrease in waiting time with the Panthera metadata cache and a 95.7\% decrease with the Panthera data cache. Such drastic decreases in latency can greatly increase the efficiency of existing Hadoop applications and also allow the creation of algorithms that were previously not feasible. Panthera has numerous applications in fields ranging from bioinformatics and medicine to market research.  All code associated with this project will be open sourced to further development in the area of distributed computing.
}

\section{Introduction}
As computational problems and their associated datasets have grown in size, it has become necessary to take advantage of distributed computing systems to solve them. Such systems allow programs to scale from one computer (i.e. one \textit{node}) to thousands of computers with few or no changes in the codebase. In particular, the Hadoop distributed system \cite{hadoop} has seen tremendous growth. Hadoop is an open source version of the revolutionary MapReduce system developed at Google \cite{mapreduce}. With it, developers can easily take advantage of large, multi-node clusters to solve computational problems.

In order to run algorithms across a network of computers, access to the dataset must be available to all nodes within a network. To facilitate this, Hadoop uses the Hadoop Distributed File System (HDFS) \cite{hdfs}, which allows one node to download files from other nodes.

Current Hadoop deployments exist in the thousands of nodes at companies such as Cloudera, Facebook, Yahoo, etc. Especially in such large networks, \textit{access latency}, or the waiting time associated with accessing files, is a significant concern. Latency in distributed systems can have significant effects, and a reduction of the same comes with tremendous benefits. As the RAMCloud project has outlined \cite{ramcloud}, low latency can greatly extend the applications of distributed computing systems. For example, tree traversal algorithms are largely impractical in Hadoop \cite{ramcloud}, however, with lower latency, such algorithms may become useful and applicable. The Hadoop File System currently does not use local (i.e. on the client) random access memory (RAM) to improve latency.

In this paper, I present \textit{Panthera}, a \textit{cache layer} for Hadoop. Panthera functions by retaining recently accessed files in RAM, thus, when the files are needed once again, they do not have to be downloaded or read from the hard drive, thereby greatly reducing file access latency. Thus, using \textit{Panthera} can save on computational resources and reduce total execution time for a large variety of applications.

In Section 2, we discuss the specific challenges associated with such a project. Section 3 details the architecture differences between a standard Hadoop installation and one with \textit{Panthera}.  Sections 4 and 5 detail the differences between metadata and data caching within \textit{Panthera}. Results and conclusions are covered in Sections 6, 7, 8, 9, concluding with a roadmap for future development in Section 10.

\section{Design Constraints}

There are several systems built on Hadoop that are in widespread use \cite{hbase, cloudbatch, pig}. The Hadoop project is also supported by large corporations with a 
consistent release cycle. Thus, for \textit{Panthera} to be practical, it must operate independently of the existing Hadoop codebase, i.e. as a layer, rather than a modification. This implies that \textit{Panthera} must implemented without a single change in the existing Hadoop codebase, i.e. as a "drop-in" solution. In order for \textit{Panthera} to remain independent of the Hadoop codebase, I had to reimplement a significant portion of the Hadoop File System network protocol, which in and of itself was an engineering challenge.

@@

\section{Metadata and Data}

Within the Hadoop File System (HDFS), there is a clear distinction between \textit{metadata} and \textit{data} \cite{hdfs}.

Metadata refers to information about a file or a directory, e.g. file size, file names, on which DataNode the contents of a file are located, etc. For HDFS, the NameNode handles all metadata. For every request, the NameNode recomputes metadata. \textit{Panthera} can reduce this computation by caching the results of a previous request, thereby conserving resources.

Data refers to the actual contents of a file. Within HDFS, each file is split into 64 MB large \textit{blocks}, which are stored on DataNodes. Here, \textit{Panthera} serves to reduce access time by holding recently used blocks in RAM locally.

Metadata and data caching provide significantly different technical challenges as the former involves conversion of computational latency to memory latency, whereas the latter involves a conversion of network and hard drive latency to memory latency. Note: memory latency is by far the lowest of all three in consideration. In many cases \cite{ramcloud} it is 1/1000th that of hard drive latency, which is the closest competitor.

@@

Figure 1 (see next page) outlines the architecture differences between a standard Hadoop File System architecture and an installation with Panthera.

@@

Currently, a standard LRU cache \cite{cache} is used for each method. Testing of different caching methods and predictive prefetching algorithms is under way.

@@

Tests for the metadata cache were conducted on both single node and multinode setups, whereas the data cache testing was run on multinode setups.

@@

\textbf{The Panthera data cache showed a 95.7\% decrease in the latency in comparison to a standard deviation non-datacached installation. In addition, the standard of deviation was lower by a factor of 27}. For any deployed project, maximum levels of latency must be considered when designing a system. Thus, the \textit{status quo} of the non-cached version severely limits possibilities within Hadoop. With \textit{Panthera}, not only is latency radically reduced, but also the variation in the latency is minimized.

@@

Metadata caching has similarly incredible results. \textbf{\textit{Panthera} showed a 9 fold or 90.8\% improvement in the total latency as measured on the client side in a multinode setup}. Additionally, the standard deviation is 24 fold lower with\textit{Panthera}. Thus, this project has been able to minimize the latency averages as well as spreads to an extremely significant degree.

@@

\textit{Panthera} fulfills the constraints mentioned at the outset. It is a layer on Hadoop rather than a main repository code modification. Considering the latency measurements, it is clear that it has also met the requirements in terms of benefit. \textit{Panthera} was able to reduce data access latency by 95.7\% and reduce basic metadata access latency by 90.8\%. Also, \textit{Panthera} was able to minimize the spread of latency times in both data and metadata requests. Thus, \textit{Panthera} opens up great possibilities within Hadoop due to the fact that algorithms that were previously impractical (e.g. distributed tree traversal algorithms) may now be implemented.

@@

I am also working on a cache-based scheduler for Hadoop. Essentially, it will intelligently use information about what files are available in various clients and arrange Hadoop tasks such that the use of the cache is maximized.
